---
title: "Lab 6 - clustering"
author: "Alessandra Puig-Santana"
date: "2023-02-16"
output: 
  html_document: 
    code_folding: show
---

```{r setup, echo = TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Attach packages
library(tidyverse)
library(janitor)
library(palmerpenguins)

## packages for cluster analysis
library(NbClust)
library(cluster)
library(factoextra)
library(dendextend)
library(ggdendro)
```

# Intro to cluster analysis - k-means, hierarchical

## Part 1: K-means clustering 
```{r}

# Exploratory plot - trying to understand the data
ggplot(penguins) +
  geom_point(aes(x= bill_length_mm,
                 y = bill_depth_mm,
                 color = species, 
                 shape = sex),
             size = 3,
             alpha = 0.7) +
  scale_color_manual(values = c('orange', 'cyan4', 'darkmagenta'))

# Exploratory plot 
ggplot(penguins) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = species, 
                 shape = sex),
             alpha = 0.7,
             size = 3) +
  scale_color_manual(values = c('orange', 'cyan4', 'darkmagenta'))

### Why isn't it the color i have chosed?
```

### Create a complete, scaled version of the data
```{r}
penguins_complete <- penguins %>%
  drop_na(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm)

penguins_scale <- penguins_complete %>%
  select(ends_with("_mm"), body_mass_g) %>%
  scale()
```

### Estimate the amonut of clusters
```{r}
number_est <- NbClust(penguins_scale, 
                      min.nc = 2, max.nc = 10, 
                      method = 'kmeans')

fviz_nbclust(penguins_scale, FUNcluster = kmeans, 
             method = 'wss', k.max = 10)
```

### run some k-means
```{r}
set.seed(123)
penguins_km <- kmeans(penguins_scale,
                      centers = 3, # number of centers we are starting with
                      iter.max = 10, 
                      nstart = 25)

# penguins_km$size
# penguins_km$cluster

# Let's create a penguins CL
penguins_cl <- penguins_complete %>%
  mutate(cluster_no = factor(penguins_km$cluster))

## ggplot 
ggplot(penguins_cl) +
  geom_point(aes(x = flipper_length_mm,
                 y = body_mass_g,
                 color = cluster_no, 
                 shape = species))

ggplot(penguins_cl) +
  geom_point(aes(x = bill_length_mm,
                 y = bill_depth_mm,
                 color = cluster_no, 
                 shape = species)) +
  scale_color_viridis_d()
```


```{r}
# Table showing the different clusters and how much was captured 
penguins_cl %>%
  select(species, cluster_no) %>%
  table()
```

### Part 2: Hierarchical clustering 

#### Start with complete linkage 
```{r}
# Create distance matrix
penguin_dist <- dist(penguins_scale, method = "euclidean")

# Hierarchical clustering (complete linkage)
penguin_hc_complete <- hclust(penguin_dist, method = "complete")
### also: single, average, word.D are other methods that return different types of clusters

### plot a dendrogram
plot(penguin_hc_complete, cex = .6, hang = -1, )

### cut dendrogram into three clusters
penguin_cut_hc <- cutree(penguin_hc_complete, 3)
table(penguin_cut_hc, penguins_complete$species)
```
## World Bank data: read in and simplify
```{r}
wb_env <- read_csv(here::here('data/wb_env.csv')) 

wb_ghg_20 <- wb_env %>%
  slice_max(n = 20, ghg) # Top 20 

summary(wb_ghg_20)

wb_scaled <- wb_ghg_20 %>%
  select(3:7) %>%
  scale()

summary(wb_scaled)

rownames(wb_scaled) <- wb_ghg_20$name
```

### Find the Ecludean distances
```{r}
euc_distance <- dist(wb_scaled, method = "euclidean")
```

### Perform hierarchical clustering with complete linkage 
```{r}
hc_complete <- hclust(euc_distance, method = "complete")
plot(hc_complete, cex = .6, hang = -1)
```

### Single linkage hierarchical clustering 
```{r}
hc_single <- hclust(euc_distance, method = "single")
plot(hc_single, cex = .6, hang = -1)
```

### Make a tanglegram (2 dendrograms combining)
```{r}
dend_complete <- as.dendrogram(hc_complete)
dend_single <- as.dendrogram(hc_single)

tanglegram(dend_complete, dend_single)

## see how messy both dendrograms are 
entanglement(dend_complete, dend_single)

# rerranging the branches to be cleaner, not changing the clusters
untangle(dend_complete, dend_single, method = "step1side") %>%
  entanglement() 

untangle(dend_complete, dend_single, method = "step1side") %>%
  tanglegram(common_subtrees_color_branches = TRUE)
```

### Let's make a dendrogram in ggplot!
```{r}
ggdendrogram(hc_complete, rotate = TRUE) +
  theme_minimal() +
  labs(x = "Country")
```

